---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# proteomicsDIA

<!-- badges: start -->
<!-- badges: end -->

The `proteomicsDIA` package will (eventually) be our internal R package that supplies all the functions we need for doing proteomics analysis of DIA data.

## Installation

The Github repository where the package is stored is private, so installation is a little more complicated that usual. You'll need the `devtools` package installed (which you'll want to have installed anyway for developing the package, see below). You'll also need the personal access token (PAT) that gives you access to the repository. Email Tim or Stephanie to get it. Once you have it, you can install the development version of `proteomicsDIA` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("ByrumLab/proteomicsDIA",
                         auth_token = "COPY_PAT_HERE")
```

## Example

At the start, `proteomicsDIA` only has one function: `uselsss_function()`:

```{r example, eval = F}
# Load the proteomicsDIA package
library(proteomicsDIA)


# call our only function
useless_function("It doesn't matter what aregument you supply")

# Get help on how to use uselss_function:
?useless_function
```

## Contributing to package development

I would highly recommend checking out the [R Packages](https://r-pkgs.org/) book from Hadley Wickham and Jenny Bryan. It is a pretty comprehensive book on creating R packages, and provides a nice framework that we can all follow (and the framework which the package is currently following). 

You'll want to install a few packages for development: 

```{r package install, eval = F}
install.packages(c("devtools", "roxygen2", "testthat", "knitr", "rmarkdown", "usethis"))
```

You may want to set up your `.Rprofile` to automatically load some of all of the packages when you start R (see [here](https://r-pkgs.org/setup.html#personal-startup-configuration) for more info), though I don't usually bother and just load them as I need them. 

### Collaboration with Git

We need to figure this out! Deciding on the best path forward with this depends a little on everyone's familiarity and level of comfort with Git. I (Tim) have used Git for the last couple years for all my projects, though I have rarely used it to collaborate with others and thus have rarely needed to worry about the optimal collaboration workflow. We can work together to figure that out more, but in the meantime here are some resources that might be helpful:

* [Happy Git with R](https://happygitwithr.com/index.html)- A guide to using Git and Github with RStudio.
* [Chapter 18 of R packages](https://r-pkgs.org/git.html) is on Git. 
* Git GUIs- Personally, I mostly avoid using Git in the command line and instead use GUIs. RStudio has a minimal one, explained in the first link above. Initially, I mostly used [GitKraken](https://www.gitkraken.com/git-client), which is free with a student license and has some nicer touches (better diff viewing, better branch visualization). Lately, I've just been using the Git functionality in [Visual Studio Code](https://code.visualstudio.com/), which also has some nice extensions for using Git. 

Once we're all set up and comfortable with Git, we'll need to figure out how to manage workflows. That might be a topic for another day. 
