---
title: "proteoDA_v2.0_design"
author: "Byrum Lab"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# One-factor design (no intercept)

```{r one_factor, eval=FALSE}
design <- ~ 0 + group
mm <- model.matrix(design, data = metadata)
colnames(mm)  # e.g., groupA, groupB, groupC
```

# Factorial design with interaction

```{r factorial_noIntercept, , eval=FALSE}
design2 <- ~ 0 + cell * treat
mm2 <- model.matrix(design2, data = metadata)
colnames(mm2)  # main + interaction terms
```

# Random block with duplicateCorrelation (v2.0 wrapper may support; otherwise do manual)

# Example sketch (pseudocode):
```{r block_dupCor, , eval=FALSE}
 dupcor <- limma::duplicateCorrelation(expr, mm, block = metadata$patient_id)
 fit <- limma::lmFit(expr, mm, block = metadata$patient_id, correlation = dupcor$consensus)
 fit2 <- limma::eBayes(limma::contrasts.fit(fit, contrasts))

```


# And a contrasts snippet aligned to a factorial:
## Suppose colnames(mm2) include cellWT, cellHET, treatSham, treatGy, and interactions

- Contrast: (HET vs WT) under Gy only:
- (cellHET + cellHET:treatGy) - (cellWT + cellWT:treatGy) simplifies to HET:Gy - WT:Gy

## In practice, build explicit contrast matrix matching colnames(mm2).
