---
title: "proteoDA v2.0 Workflow: DIANN → proteoDA (with/without Perseus imputation)"
author: "Byrum Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r knitsetup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=TRUE, warning = TRUE, fig.width=7, fig.height = 5)
```

# Overview

This vignette demonstrates the `proteoDA version 2.0` workflow using `DIANN` protein quantification as input. It runs the analysis twice:

1. Without imputation (leaving NA where values are missing).
2. With Perseus-style imputation (downshifted Gaussian).

## Outputs include:

A) Per-contrast limma results with z-score based effect size standardization.
B) Combined results tables and an Excel workbook with per-contrast sheets.
C) Interactive volcano/report HTMLs per contrast.
D) Perseus imputation diagnostics (before/after distributions).
E) Saved DAList objects for downstream QC and reporting.

## Requirements

1. `proteoDA >= 2.0`
2. A project `proteoDA_params.R` defining paths/labels/thresholds, e.g.:
  
  A. Define input and output file paths using: `input_quan`, `metadata`, `contrasts`, `QC_dir`, and `DA_dir` 
  B. Define annotation and sample start/end columns in `input_quan` using `anno_start`, `anno_end` and `sample_start` 
  C. Define limma design formula with `design`
  D. Define sample group information with `group` 
  E. Define protein filtering criteria with `filt_min_reps`, `filt_min_groups`, and `require_both_groups`
  F. Define statistical significance thresholds with `p.val`, `logFC`, `bin_size`, `stat_cols`
  G. Define visualization and reporting columns with `DA_table_cols`, `DA_title_col`, `ctrl_proteins`, `project_name`, `author`

**More details:**
- Data preprocessing (import, alignment, filtering): see `vignette("proteoDA_v2.0_preprocessing")`
- Normalization & imputation: see `vignette("proteoDA_v2.0_Norm_Impute")`
- Designs & contrasts (basic → advanced): see `vignette("proteoDA_v2.0_design")`

# Installation

To install proteoDA version 2.0 from GitHub (with dependencies and vignettes built):

```{r install, eval=FALSE}
devtools::install_github("ByrumLab/proteoDA2.0",
                         dependencies = TRUE,
                         build_vignettes = TRUE,
                         force = TRUE)

```

# Load Packages and Parameters

```{r load_package}
library(proteoDAstjude)
library(yaml) # used to generate the PowerPoint summary

#browseVignettes(package = "proteoDAstjude")

# Load your project parameters (defines inputs/labels/thresholds)
source("P:/projects/CPM_publications/03_proteoDA2_Steph/code/proteoDA_params.R")
```

# Data setup to create a proteoDA `DAList` object for analysis

`proteoDA` ships with small DIA-style test datasets for quick evaluation.

[add example data to package]

```{r package_data}
# Read DIANN quant table (CSV)

stopifnot(file.exists(input_quan))
input_data <- read.csv(input_quan, stringsAsFactors = FALSE, check.names = FALSE)

# Defensive checks

if (!("Protein.Group" %in% colnames(input_data))) {
stop("Input data must contain a 'Protein.Group' column.")
}
```

## Make `uniprot_id` and Clean Missing/Blank Entries

```{r 3_create_uniprot_id}
# Extract first UniProt ID from semicolon-delimited Protein.Group
uniprot_split <- strsplit(input_data$Protein.Group, ";", fixed = TRUE)
uniprot_id <- vapply(uniprot_split, function(x) if (length(x)) x[[1]] else NA_character_, "")
input_data <- cbind(uniprot_id = uniprot_id, input_data)

# Replace NAs and blanks: numeric -> 0; character -> "0"
for (nm in names(input_data)) {
  col <- input_data[[nm]]
if (is.numeric(col)) {
  col[is.na(col)] <- 0
  input_data[[nm]] <- col
} else {
  col[is.na(col) | col == ""] <- "0"
  input_data[[nm]] <- col
}
}

```

## Load Sample Metadata and Protein Annotation 

```{r 3_metadata_anno_load}
# Load sample metadata
#stopifnot(file.exists(metadata))
sample_metadata <- read.csv(metadata, stringsAsFactors = FALSE, check.names = FALSE)

# Protein annotation slice from input file using anno_start:anno_end from params
if (!exists("anno_start") || !exists("anno_end")) {
  stop("Parameters 'anno_start' and 'anno_end' must be defined in proteoDA_params.R")
}
annotation_data <- input_data[, anno_start:anno_end, drop = FALSE]

```

## Check sample metadata and the `input_quan` samples match

### Verify the sample order in both the sample metadata and the protein intensity columns are named the same and in the same order. If not the `align_data_metadata()` will correct the data frames so the sample group information is correct. 

```{r 3_check_samples}

# Rename data columns: metadata$file -> metadata$sample where present

req_cols <- c("file", "sample")
missing_req <- req_cols[!(req_cols %in% colnames(sample_metadata))]
if (length(missing_req)) {
stop(paste("Metadata must contain columns:", paste(req_cols, collapse = ", ")))
}
rename_map <- setNames(sample_metadata$sample, sample_metadata$file)
present_files <- intersect(names(rename_map), colnames(input_data))

# Apply renaming in-place

if (length(present_files)) {
idx <- match(present_files, colnames(input_data))
colnames(input_data)[idx] <- rename_map[present_files]
}

# Build intensity matrix keeping only samples present in metadata

dat_ids <- colnames(input_data)
keep_samples <- dat_ids[dat_ids %in% sample_metadata$sample]
intensity_data <- input_data[, keep_samples, drop = FALSE]

# Trim metadata to these samples; rownames as sample IDs

sample_metadata <- sample_metadata[sample_metadata$sample %in% keep_samples, , drop = FALSE]
rownames(sample_metadata) <- sample_metadata$sample

# Auto-align (reorders and checks)

aligned <- align_data_and_metadata(
data                = intensity_data,
metadata            = sample_metadata,
sample_col          = "sample",
group_col           = "group",        # optional but recommended
prefer_group_blocks = FALSE,          # set TRUE to keep group blocks compact
strict              = FALSE
)

# (Optional) show alignment changes

aligned$changes

# Replace with aligned data

intensity_data  <- aligned$data
sample_metadata <- aligned$metadata

```

# Create `DAList` for analysis in proteoDA 2.0  

```{r 4_raw_DAList}
raw <- DAList(
data      = intensity_data,
annotation= annotation_data,
metadata  = sample_metadata,
design    = NULL,
eBayes_fit= NULL,
results   = NULL,
tags      = NULL
)

# Convert embedded zeros to NAs for downstream missingness-aware steps
filtered_samples <- zero_to_missing(raw)
# Quick sanity: group counts
table(filtered_samples$metadata$group)

```

# Protein Filtering 

A new function `filter_proteins_per_contrast()` has been added to separately filter each sample group comparison in the `contrasts.csv` based on the user selected thresholds. The `filter_proteins_by_group()` will filter all sample groups together for a more global filtering approach (see vignette/proteoDA_v1.0_workflow.Rmd). This may all proteins that are missing in two sample groups to still be present in the data for that particular limma analysis. 

## Filter_proteins_per_contrast()

### Requires a `contrasts.CSV` defining sample group comparisons.

```{r 5_protein_filtering}

stopifnot(file.exists(contrasts))

filtered_DALists <- filter_proteins_per_contrast(
  DAList              = filtered_samples,
  contrasts_file      = contrasts,
  min_reps            = filt_min_reps,
  require_both_groups = require_both_groups,
  grouping_column     = group
)

# Summarize counts
summary_df <- data.frame(
  contrast = names(filtered_DALists$filtered_proteins_per_contrast),
  filtered_proteins = vapply(filtered_DALists$filtered_proteins_per_contrast, length, 1L)
)
write.csv(summary_df, "filtered_protein_counts.csv", row.names = FALSE)

```

# Normalization Methods

`proteoDA` includes 8 common normalization methods including: log2 transformation (`log2`), log2 median (`median`), log2 mean (`mean`), variance stablizing normalization (`vsn`), log2 quantile (`quantile`), cyclic loess (`cycloess`), robust linear regression (`rlr`), and log2 global intensity (`gi`). 

If the data has been normalized using an outside method, then proteoDA normalization can be skipped. However, the normalization method must be defined in the `DAList$tag$norm_method` slot for reporting. 

```{r 6_normalization}
# Using DIANN's log2 median normalized values directly

# copies the DAList first and then adds the norm method tag
norm <- filtered_DALists
norm$tags$norm_method <- "diann_quan"

```

# Optional: Perseus-Style Imputation using the normal distribution

## proteoDA 2.0 now includes the option to impute missing values using the downshifted Gaussian distribution similar to that of the Perseus software by Cox lab. 

**Reference:** Tyanova, Stefka, and Juergen Cox. 2018. “Perseus: A Bioinformatics Platform for Integrative Analysis of Proteomics Data in Cancer Research.” In, 133–48. Springer New York. https://doi.org/10.1007/978-1-4939-7493-1_7. 

```{r 7_Perseus_imputation}

# Perseus’s normal distribution imputation (downshifted Gaussian on normalized log2)
set.seed(1)
imputed <- perseus_impute(
    norm,
    shift = 1.8,
    width = 0.3,
    robust = TRUE,             # logical, use median/MAD (TRUE) or mean/SD (FALSE) per sample
    save_before_after = TRUE,  # used to generate histogram figure
    seed = 1
)

# Diagnostic plots (histograms/densities before vs after), saved if out_dir provided

plots <- write_perseus_imputation_plots(
    DAList      = imputed,
    out_dir     = "PerseusPlots",
    bins        = 30,
    facet_ncol  = 4,
    overlay     = TRUE,
    width       = 7,
    height      = 7,
    dpi         = 300,
    device      = "png"
)

# To preview one plot in an interactive R session:
# print(plots[["YOUR_CONTRAST_NAME"]])

```

# Design Sanity Check (Full Rank & Aliasing) 

```{r 8_design_check, eval=FALSE}
# Build a model frame from metadata according to your 'design' formula
if (!exists("design")) stop("Parameter 'design' must be defined (e.g., ~0 + group).")

mf <- model.frame(formula = design, data = norm$metadata, drop.unused.levels = TRUE)
mm <- model.matrix(attr(mf, "terms"), data = mf)

# 1) Full rank check
qr_rank <- qr(mm)$rank
is_full_rank <- (qr_rank == ncol(mm))

cat("Design matrix columns:", ncol(mm), "\n")
cat("Design matrix rank   :", qr_rank, "\n")
cat("Full rank?           :", is_full_rank, "\n\n")

# 2) Aliased coefficients (if any)

aliased <- alias(lm.fit(x = mm, y = rnorm(nrow(mm))))$Complete

if (!is.null(aliased)) {
  cat("Aliased (non-estimable) coefficient relationships detected:\n")
  print(aliased)
} else {
  cat("No aliased coefficients detected.\n")
}


```

## Troubleshooting: Design Matrix and Contrast Mismatch

If you see an error in `run_filtered_limma_analysis()` such as:

- Error in contrasts.fit(...):
- row names of contrast matrix do not match column names of design matrix

this means the **contrast file labels** (e.g., group names or coefficients) don’t match the **design matrix column names**.  
Use the following diagnostic commands to inspect and correct them.

```{r troubleshoot_design, eval=FALSE}
# ---- Inspect design and contrasts ----
# Show design matrix column names

cat("Design matrix columns:\n")
print(colnames(mm))

# Read and inspect your contrasts CSV
cat("\nDefined contrasts from file:\n")
contrasts_df <- read.csv(contrasts, stringsAsFactors = FALSE)
print(contrasts_df)

# Example: If your design formula was ~0 + group
# then design matrix columns should look like "A", "B", etc.
# Ensure that the contrasts file refers to those names exactly.

# Check for mismatched column names
bad_cols <- setdiff(unique(unlist(contrasts_df)), colnames(mm))
if (length(bad_cols)) {
  cat("\n⚠️ The following contrast terms do not match design columns:\n")
  print(bad_cols)
  cat("\nPlease correct your contrasts.csv to use the exact column names above.\n")
} else {
  cat("\n✅ All contrast terms match design matrix columns.\n")
}

# Optional: visualize design matrix
head(mm)

```

### 💡 If you still encounter errors

You can build your design and contrast matrices manually using the legacy version 1.0 syntax from the original `proteoDA`:

``` {r proteoDA1.0_design, eval = FALSE}
# Recreate design and contrast matrices explicitly
DAList <- add_design(DAList = norm, design_formula = ~0 + group)
DAList <- add_contrasts(DAList = DAList, contrasts_file = contrasts)
DAList <- fit_limma_model(DAList)
```

This will let you see the internal column naming directly before running
`run_filtered_limma_analysis()` in version 2.0.

If the design works with the old commands but fails in v2.0,
check for renamed groups, factor levels, or spaces in your contrast file headers.

**Summary of what this adds:**
- Shows users how to print and compare `colnames(mm)` vs. contrast terms.  
- Warns when contrasts don’t match design.  
- Suggests reverting to `add_design()`, `add_contrasts()`, and `fit_limma_model()` if troubleshooting deeper issues.  
 

# Run Limma (No Imputation vs Perseus Imputation) 

```{r 9_run_limma}
# Common args

common_args <- list(
design_formula  = design,
pval_thresh     = p.val,
lfc_thresh      = logFC,
adj_method      = "BH",
binsize         = bin_size,
plot_movingSD   = TRUE,
contrasts_file  = contrasts,
plot_movingSD   = FALSE
)

# 1) Without imputation (leave NAs)
results <- do.call(run_filtered_limma_analysis, c(list(DAList = norm), common_args))

# 2) With Perseus imputation
results2 <- do.call(run_filtered_limma_analysis, c(list(DAList = imputed), common_args))

```

# Build Statlists, Write Tables, and generate plots

## Write Tables

```{r 10_statlists, eval = FALSE}
# Build "full universe" statlists across all contrasts for each run

statlist  <- build_statlist(DAList = results,  stat_cols = stat_cols)
statlist2 <- build_statlist(DAList = results2, stat_cols = stat_cols)

# Write per-contrast CSVs + combined CSV + Excel with sheets

dir.create(DA_dir, showWarnings = FALSE, recursive = TRUE)

write_limma_tables(
    results,                     # A DAList object with statistical results.
    output_dir         = DA_dir, # Directory to output tables. Defaults to working directory.
    overwrite          = TRUE,   # Logical. Overwrite existing files?
    contrasts_subdir   = NULL,   # Subdirectory for per-contrast CSV files
    summary_csv        = NULL,   # Filename for summary CSV
    combined_file_csv  = NULL,   # Filename for combined results CSV.
    spreadsheet_xlsx   = NULL,   # Filename for Excel spreadsheet
    add_filter         = TRUE,   # Logical. Add filters to Excel columns?
    color_palette      = NULL,   # Optional color palette for Excel output.
    add_contrast_sheets= TRUE,   # Logical. Whether to add each per-contrast CSV as a worksheet in the Excel file.
    statlist           = statlist # Optional list of per-contrast result tables to use instead of DAList$results
)

## --- Imputed results -------
dir.create("impute_out", showWarnings = FALSE, recursive = TRUE)

write_limma_tables(
    results2,
    output_dir         = "impute_out",
    overwrite          = TRUE,
    contrasts_subdir   = NULL,
    summary_csv        = NULL,
    combined_file_csv  = NULL,
    spreadsheet_xlsx   = NULL,
    add_filter         = TRUE,
    color_palette      = NULL,
    add_contrast_sheets= TRUE,
    statlist           = statlist2
)


```

## Write Plots

### These functions produce a standalone interactive HTML reports per contrast and static images

```{r 10_write_plots, eval=FALSE}
# no imputation
write_limma_plots(
    DAList           = results,
    grouping_column  = group,
    table_columns    = DA_table_cols, # values must be present in protein annotation
    title_column     = DA_title_col,  # value displayed in dotplots in interactive htmls
    height           = 1000,
    width            = 1000,
    output_dir       = DA_dir,
    overwrite        = TRUE,
    control_proteins = ctrl_proteins, # proteins of known biological interest to highlight in volcanos
    highlight_by     = "uniprot_id",  
    image_formats    = c("pdf","png")
)

# Perseus imputed data 
write_limma_plots(
    DAList           = results2,
    grouping_column  = group,
    table_columns    = DA_table_cols,
    title_column     = DA_title_col,
    height           = 1000,
    width            = 1000,
    output_dir       = "impute_out",
    overwrite        = TRUE,
    control_proteins = ctrl_proteins,
    highlight_by     = "uniprot_id",
    image_formats    = c("pdf","png")
)

```

You’ve completed the **proteoDA v2.0** workflow!

- **Results without imputation** are saved under `DA_dir/`
- **Results with Perseus imputation** are saved under `impute_out/`
- **Interactive per-contrast HTML reports** are created in each folder  
  (open them directly in your web browser)
- **Combined results tables** and **Excel workbooks** are written automatically
- The final `.rds` files are saved in your working directory:
  - `results_NA.rds` → with `NA`s retained  
  - `results_imputed.rds` → after Perseus imputation  
  - `results.rds` → zero-filled version for QC/PCA utilities
- Session details are logged in `sessionInfo.txt`
- All project variables are updated in `_variables.yml`

---

💡 **Next steps:**
- Explore significant hits using volcano or PCA plots  
- Integrate results into pathway or enrichment analysis  
- Reuse the saved `DAList` objects in downstream QC and report generation

## Notes & Tips

1. Ensure your contrast file and metadata agree with your design. The diagnostic chunk reports full rank and aliasing issues early.

2. If your metadata–data sample IDs were out of order (e.g., lexicographic 1, 10, 11, 2…), align_data_and_metadata() fixes this using name matching, not position.

3. If the DIANN export has different column names or annotation blocks, adjust anno_start:anno_end accordingly.
