---
title: "proteoDA v2.0 Preprocessing"
author: "Byrum Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    number_sections: true
vignette: >
  %\VignetteIndexEntry{proteoDA v2.0 Preprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Introduction

Preprocessing is a critical first step in any proteomics differential abundance (DA) workflow.
In `proteoDA`, we separate preprocessing (cleaning and filtering) from modeling (limma-based statistical analysis) so that you can:

- Explicitly document which samples and proteins were removed and why
- Apply different filtering rules for all groups vs contrast-specific analyses
- Track preprocessing decisions via the DAList$tags slot for reproducibility

This vignette focuses on preprocessing functions that operate on a `DAList`:

- Converting embedded zeros to missing values
- Filtering samples using metadata
- Filtering proteins using annotation
- Filtering proteins using missingness patterns across groups
- Filtering proteins per contrast using a contrasts file

Normalization, imputation, and limma designs/statistical thresholds are covered in separate vignettes:

- **Overall workflow:** `vignette("proteoDA_v2.0_workflow", package = "proteoDA")`
- **Normalization and imputation:** `vignette("proteoDA_v2.0_Norm_Impute", package = "proteoDA")`
- **Limma designs and statistical thresholds:** `vignette("proteoDA_v2.0_design", package = "proteoDA")`

Typical `proteoDA` preprocessing flow:

1. Import data → create a `DAList`
2. Convert zeros → `zero_to_missing()`
3. Optional sample-level and annotation-based filtering (`filter_samples()`, `filter_proteins_by_annotation()`)
4. Protein-level missingness filtering (`filter_proteins_by_group()` or `filter_proteins_by_proportion()`)
5. Optional contrast-specific filtering (`filter_proteins_per_contrast()`)
6. Normalize / impute → fit limma models 
 See the normalization vignette for details on available normalization strategies and diagnostics:
 `vignette("proteoDA_v2.0_Norm_Impute", package = "proteoDA")`
7. Write tables and plots → `write_limma_tables()` and `write_limma_plots()`

***
# Overview of preprocessing functions

## Summary table

```{r 2_filtering_table, echo = FALSE}
opts <- data.frame(
Function = c(
"zero_to_missing",
"missing_to_zero",
"filter_samples",
"filter_proteins_by_annotation",
"filter_proteins_by_group",
"filter_proteins_by_proportion",
"filter_proteins_per_contrast"
),
Purpose  = c(
"Convert embedded zeros to NA for missing-aware downstream steps",
"Convert NA back to 0 for export or external tools",
"Remove samples based on metadata conditions (e.g., QC failures)",
"Remove proteins based on annotation (e.g., contaminants, keratins)",
"Global missingness filtering across groups using replicate counts",
"Global missingness filtering across groups using proportions",
"Contrast-specific filtering using non-missing counts in groups"
),
Key_Args = c(
"-",
"-",
"condition (expression on metadata columns)",
"condition (expression on annotation columns)",
"min_reps, min_groups, grouping_column",
"min_prop, grouping_column",
"min_reps, require_both_groups, grouping_column, contrasts_file"
),
When_to_use = c(
"Early: almost always, before filtering and modeling",
"Late: before exporting to tools that require zeros",
"When specific samples should be dropped (e.g., outliers, failed runs)",
"When specific proteins should be dropped (e.g., keratins, decoys, spikes)",
"When all groups should share a common missingness rule",
"When group sizes are imbalanced but you want comparable missingness per group",
"When different contrasts need different missingness rules"
),
Output = c(
"DAList with NAs representing missingness",
"DAList with zeros instead of NAs",
"DAList with reduced sample set",
"DAList with reduced protein set",
"DAList (globally filtered)",
"DAList (globally filtered)",
"DAList with per-contrast filtered protein sets and data_per_contrast"
),
stringsAsFactors = FALSE
)

knitr::kable(opts, caption = "Preprocessing options in proteoDA v2.0")

```

## General usage patterns

All preprocessing functions accept and return a `DAList`, allowing you to chain them:

```{r 2_usage, eval=FALSE}
library(proteoDA)

DAList_clean <- DAList_raw |>
            zero_to_missing() |>
            filter_samples(batch != "bad_batch") |>
            filter_proteins_by_annotation(!grepl("KERATIN", protein_name)) |>
            filter_proteins_by_group(
                min_reps       = 2,
                min_groups     = 2,
                grouping_column = "group"
)
```

Most functions also update `DAList$tags` so decisions can be inspected later.

***

# Handling zeros and missing values

## zero_to_missing()

Embedded zeros often represent missing intensities rather than true zero abundance.
Statistical modeling in `proteoDA` assumes missingness is encoded as `NA`.

Typical usage:

```{r 3_zerotoNA, eval=FALSE}
DAList <- zero_to_missing(DAList)
```

When to use

- Early in the pipeline, before filtering by missingness or running limma
- Any time upstream tools encode absence as 0 rather than NA

What it does

- Replaces 0 values in `DAList$data` with NA
- Leaves metadata and annotation unchanged
- Optionally records a tag in `DAList$tags$zero_to_missing` (implementation specific)

## missing_to_zero()

In some cases you may need to convert NA back to 0, e.g., when:

- Exporting to tools that cannot handle NA, such as PCA. 
- Re-using data in pipelines that assume zeros

Typical usage:

```{r 3_missingtozero, eval=FALSE}
DAList_export <- missing_to_zero(DAList)
```

You should not run limma on zero-filled data; use zeros only for export or visualization where appropriate.

***

# Sample-level filtering
## filter_samples(): drop samples using metadata

`filter_samples()` removes samples from a `DAList` based on logical conditions evaluated on `DAList$metadata`.

Function recap

```{r 4_filter_samples, eval=FALSE}
DAList <- filter_samples(DAList,
                  condition)
```

- condition is an expression using columns from DAList$metadata
- Samples are kept where condition evaluates to TRUE

Example: remove a specific sample

```{r 4_filtsample, eval=FALSE}
DAList <- filter_samples(DAList,
                  sample_ID != "Sample_001")
```

Example: keep only "control" and "treated" groups

```{r 4_filtKeep, eval=FALSE}
DAList <- filter_samples(DAList, 
                         group %in% c("Control", "Treated"))
```

Example: remove all samples in a group

```{r 4_filtRemoveGroup, eval=FALSE}
DAList <- filter_samples(DAList,
                         group != "Pool")
```

Example: remove pre-2018 samples

```{r 4_filtYear, eval=FALSE}
DAList <- filter_samples(DAList,
                      year >= 2018)

```

Behavior and messages

- Updates `DAList$metadata` and `DAList$data` to keep only selected samples
(`DAList$data` columns are re-subset to the remaining row names in `metadata`)

- Prints how many samples were removed and shows the removed metadata
- Errors if the filter produces any NA values (to avoid silent mis-filtering)

***

# Protein-level filtering using annotation
## filter_proteins_by_annotation()

`filter_proteins_by_annotation()` removes proteins using logical conditions on `DAList$annotation`.
This is useful for removing contaminants, such as keratins, decoys, spike-in standards, etc.

Function recap

```{r 5_byAnnotation, eval=FALSE}
# general command
DAList <- filter_proteins_by_annotation(DAList,
            condition)
```

- `condition` is evaluated in the context of `DAList$annotation`
- Proteins are kept where condition is `TRUE`

Example: remove a single protein

```{r 5_singleProtein, eval = FALSE}
DAList <- filter_proteins_by_annotation(DAList,
            uniprot_id != "P12345")
```

Example: remove keratins and contaminants

```{r 5_keratin, eval = FALSE}
# Remove any protein which contains "keratin" in the name:
DAList <- DAList |>
      filter_proteins_by_annotation(!grepl("KERATIN", protein_name)) |>
      filter_proteins_by_annotation(contaminant_flag == FALSE)
```

Example: remove proteins with molecular weight < 30 kDa

```{r 5_MW, eval = FALSE}
# Remove any protein with molecular weight < 30 kDa
DAList <- filter_proteins_by_annotation(DAList,
                                          molecular_weight > 30)
```

Behavior and tags

- Updates `DAList$annotation` and re-subsets `DAList$data` rows
- On success, `DAList$tags$filter_proteins_by_annotation` is updated with the condition call
- Errors if the condition evaluates to `NA` for any row

***

# Protein-level filtering by missingness across groups

For DA analysis, you typically require some minimum level of non-missing data per group.
`proteoDA` provides two global strategies:

- By `counts` (number of replicates in number of groups) → `filter_proteins_by_group()`
- By `proportion` (fraction of samples per group) → `filter_proteins_by_proportion()`

Both depend on a grouping column in `DAList$metadata` (e.g. "group").

## `filter_proteins_by_group()`: numeric replicate thresholds

**General Rule of Thumb:** Require at least 2/3 of biological replicates to have values in at least one group. 

Limma requires at least 2 replicates but power increases with increased replicates due to better estimation of model variance. 

For each protein:

- Count how many samples per group have non-missing values
- Require at least `min_reps` replicates within at least `min_groups` groups

Function recap

```{r 6_bygroup, eval=FALSE}

DAList_filtered <- filter_proteins_by_group(
                DAList,
                min_reps       = 2,
                min_groups     = 1,
                grouping_column = "group"
)

```

Arguments

- `min_reps` (required) \
Minimum number of non-missing samples within a group for a protein to be considered present

- `min_groups` (required) \
Minimum number of groups that must meet min_reps

- `grouping_column` \
Column in `DAList$metadata` with group labels (default "group")

Example: strict filtering (no missingness allowed)

- data contains 5 biological replicates and 4 conditions or "groups"

```{r 6_strict, eval=FALSE}

DAList_strict <- filter_proteins_by_group(
          DAList,
          min_reps       = 5,     # group size
          min_groups     = 4,     # all groups
          grouping_column = "group"
)
```

Example: at least one replicate in all groups

```{r 6_lax, eval=FALSE}
DAList_lax <- filter_proteins_by_group(
              DAList,
              min_reps       = 1,
              min_groups     = 4,
              grouping_column = "group"
)

```

Behavior and diagnostics

- Checks that `min_reps` is not larger than the smallest group size \
(otherwise errors with an informative message)
- Checks that enough `groups` can meet `min_reps` to satisfy `min_groups` 
- Builds an internal matrix of non-missing counts per protein × group
- Returns a filtered `DAList` and logs parameters in `DAList$tags$filter_proteins_by_group`

## filter_proteins_by_proportion(): proportional thresholds

When group sizes are imbalanced, you may prefer to specify a proportion rather than an absolute number of replicates.

For each protein and group:

- Compute group-specific thresholds = ceiling(group_size * min_prop)
- Require that the protein is present in at least threshold samples in each group

Function recap

```{r 6_proportion, eval=FALSE}
DAList_prop <- filter_proteins_by_proportion(
            DAList,
            min_prop       = 0.66,
            grouping_column = "group"
)

```

Arguments

- `min_prop` \
Minimum fraction of samples within each group that must be non-missing \
(must be between 0 and 1, e.g. 0.5, 0.66, 1.0)

- `grouping_column` \
Column in `DAList$metadata` with group labels

Example: strict (no missing)

```{r 6_strictProportion, eval=FALSE}
DAList_all_present <- filter_proteins_by_proportion(
        DAList,
        min_prop       = 1,
        grouping_column = "group"
)
```

Example: moderate (≥50% of samples per group)

```{r 6_modProp, eval=FALSE}
DAList_half <- filter_proteins_by_proportion(
      DAList,
      min_prop       = 0.5,
      grouping_column = "group"
)
```

Behavior and tags

- Computes group-specific thresholds via ceiling(n_group * min_prop)
- Requires all groups to meet their thresholds for a protein to be kept
- Updates `DAList$data`, `DAList$annotation`, and logs parameters in `DAList$tags$filter_proteins_by_proportion`

***

# Contrast-specific missingness filtering

Global filters may be too strict or too lenient for some contrasts, especially in heterogeneous designs.

`filter_proteins_per_contrast()` allows you to:

- Use the same raw `DAList` for all contrasts
- Apply `per-contrast` missingness rules based on the groups involved in each contrast
- Store results in `DAList$data_per_contrast` and `DAList$annotation_per_contrast` for downstream limma runs

## **filter_proteins_per_contrast()**
Concept

Given:

- A `contrasts_file` where each row looks like `GroupA_vs_GroupB=GroupA - GroupB`
- A `grouping_column` in metadata (e.g., "group")

For each contrast:

- Identify the two groups in the contrast (e.g., "GroupA" and "GroupB")
- Count non-missing values per protein in each group
- Apply `min_reps` and `require_both_groups` to decide which proteins to keep
- Store contrast-specific data and annotation

Function recap

```{r 8_percontrast, eval=FALSE}
DAList_filtered <- filter_proteins_per_contrast(
        DAList           = DAList,
        contrasts_file   = "data/contrasts.csv",
        min_reps         = 2,
        require_both_groups = TRUE,
        grouping_column  = "group"
)
```

Arguments

- `DAList` \
Input `DAList` with raw or pre-filtered data, metadata, annotation

- `contrasts_file` \
Path to CSV with lines like `GroupA_vs_GroupB=GroupA - GroupB` (no header)

- `min_reps` \
Minimum replicates required per group in the contrast

- `require_both_groups` \
**TRUE:** both groups must meet `min_reps` \
**FALSE:** at least one group must meet `min_reps`

- `grouping_column` \
Column in `metadata` with `group` labels

Outputs and bookkeeping

On success, the function:

- Adds `DAList$filtered_proteins_per_contrast`
  (named list of retained IDs per contrast)

- Adds `DAList$data_per_contrast`
  (named list of contrast-specific data matrices)

- Adds `DAList$annotation_per_contrast`
  (named list of contrast-specific annotation data frames)

- Adds `DAList$tags$retention_summary`
  (data frame with Contrast and Proteins_Retained)

These per-contrast slots are then used by downstream functions such as `run_filtered_limma_analysis()` or plotting/reporting functions that operate per contrast.

*NOTE:* If you use `filtered_proteins_per_contrast()` and then apply `normalize_data()`, the normalized values will be added to `DAList$data_per_contrast` and `DAList$data` will remain untouched. 

Example: heterogeneous design

```{r 8_heterogeneou, eval=FALSE}
DAList <- zero_to_missing(DAList)

DAList <- filter_proteins_per_contrast(
        DAList             = DAList,
        contrasts_file     = "design/contrasts.csv",
        min_reps           = 2,
        require_both_groups = FALSE,
        grouping_column    = "group"
)

DAList$tags$retention_summary

```


***

# Example workflows

## Simple balanced design


```{r 8_exSimple, eval=FALSE}
DAList_pp <- DAList_raw |>
        zero_to_missing() |>
        filter_samples(group %in% c("Control", "Treatment")) |>
        filter_proteins_by_annotation(!grepl("KERATIN", protein_name)) |>
        filter_proteins_by_group(
            min_reps       = 2,
            min_groups     = 2,
            grouping_column = "group"
)

```

## Imbalanced groups with many samples

```{r 8_exImbalanced, eval=FALSE}
DAList_pp <- DAList_raw |>
    zero_to_missing() |>
    filter_samples(!qc_fail) |>
    filter_proteins_by_proportion(
        min_prop       = 0.5,
        grouping_column = "group"
)
```

## Mixed design with contrast-specific filters

```{r 8_exContrastSpecific, eval=FALSE}
DAList_pp <- DAList_raw |>
    zero_to_missing() |>
    filter_proteins_by_annotation(!decoy) |>
    filter_proteins_per_contrast(
        DAList             = _,
        contrasts_file     = "data/contrasts.csv",
        min_reps           = 2,
        require_both_groups = FALSE,
        grouping_column    = "group"
)

```

***

# Summary

In `proteoDA`, preprocessing is a separate, explicit step designed to make downstream limma modeling more robust and more transparent. The key ideas from this vignette are:

- **Handle zeros early**  
  Use `zero_to_missing()` so that missingness is consistently encoded as `NA` before any filtering or modeling.

- **Filter at the sample level when needed**  
  Use `filter_samples()` with metadata-based conditions to remove poor-quality, failed, or out-of-scope samples before touching protein-level filters.

- **Use annotation to remove unwanted proteins**  
  Use `filter_proteins_by_annotation()` to remove contaminants, keratins, decoys, spike-ins, or any proteins flagged by annotation.

- **Control missingness across groups**  
  - `filter_proteins_by_group()` for replicate-count–based thresholds (`min_reps`, `min_groups`)  
  - `filter_proteins_by_proportion()` when group sizes are imbalanced and a proportion-based rule (`min_prop`) is easier to reason about.

- **Apply contrast-specific rules when designs are heterogeneous**  
  `filter_proteins_per_contrast()` allows you to tailor missingness thresholds to each contrast and stores per-contrast data and annotation in `DAList$data_per_contrast` and `DAList$annotation_per_contrast`.

- **Track decisions via tags**  
  Most preprocessing functions write into `DAList$tags`, making it easier to document which filters were applied and with what parameters.

Together, these tools let you build a reproducible preprocessing pipeline that matches your experimental design, balances sensitivity and robustness, and feeds clean, well-documented inputs into the `proteoDA` modeling and reporting functions.

***

# FAQ
## Should I always run `zero_to_missing()?`

In most DIA/DDA proteomics workflows, yes:

- Upstream tools often encode non-detected values as 0, NA, or NaN
- proteoDA has a function in the beginning of the workflow that finds blanks and NAs and replaces those with a numeric `0` so it is clear what input data values are included
- Downstream modeling in `proteoDA` assumes missing values are `NA`
- Running `zero_to_missing()` early ensures consistent missing value handling

The only time you might not need it is if your input data are already NA-coded or you want to run limma using zero values specifically.

***

## When should I filter samples vs proteins?

Use `filter_samples()` for:

- Poor-quality or flagged samples
- Known batch failures
- Removing specific sample subgroups

Use *protein-level* filters (`filter_proteins_by_annotation`, `filter_proteins_by_group`, `filter_proteins_by_proportion`, `filter_proteins_per_contrast`) for:

- Contaminants, decoys, spike-ins
- Insufficient data density (missingness)

In practice, you typically:

1. Filter samples first
2. Then apply protein-level filters based on the cleaned sample set

***

## How strict should my missingness filters be?

There is no single rule, but some common patterns:

- Exploratory analysis
  - filter_proteins_by_group(min_reps = 1, min_groups = n_groups) \
or \
  - filter_proteins_by_proportion(min_prop = 0.5) \

- Confirmatory analysis / publication-ready
  - filter_proteins_by_group(min_reps = 2 or 3, min_groups >= 2)
  - Possibly combine with imputation restricted to filtered proteins

We recommend trying a few thresholds and checking how many proteins remain and how model diagnostics look (e.g., moving SD plots).

***

## When should I use `filter_proteins_per_contrast()`?

Use contrast-specific filtering when:
- Your design includes heterogeneous groups or many unused samples for a given contrast
- Some contrasts have fewer replicates per group and need a slightly different rule
- Some groups have more missing values
- You want to maximize proteins per contrast without globally over-filtering

If your experiment is simple and balanced, global filters may be enough. The QC missing value heatmap can help decide how much filtering is required. 

***

## How do preprocessing choices interact with normalization and imputation?

A typical (recommended) order:

1. Import → create DAList
2. zero_to_missing()
3. Sample-level and annotation-level filtering
4. Protein-level missingness filtering (filter_proteins_by_group / filter_proteins_by_proportion / filter_proteins_per_contrast)
5. Normalization (see normalization vignette)
6. Imputation (if desired, after normalization)
7. Fit limma models (see design/statistics vignettes)

See the normalization vignette for details on available normalization strategies and diagnostics:
`vignette("proteoDA_v2.0_Norm_Impute", package = "proteoDA")`

The choice of limma design and contrasts is described in more detail in:
`vignette("proteoDA_v2.0_design", package = "proteoDA")`


Consistency is key: once you decide on a pipeline, keep the same sequence for all datasets and document thresholds in the project reports.

***

## How can I check what filtering was applied?

Many preprocessing functions store parameters under `DAList$tags`. For example:

```{r 10_FAQtags, eval=FALSE}
str(DAList$tags, max.level = 1)

DAList$tags$filter_proteins_by_group

DAList$tags$filter_proteins_by_proportion

DAList$tags$retention_summary
```

You can also directly inspect the dimensions of the data and annotation:

```{r 10_FAQdim, eval=FALSE}
dim(DAList$data)

dim(DAList$annotation)
```

For contrast-specific filtering:

```{r 10_FAQsummary, eval=FALSE}
names(DAList$data_per_contrast)

DAList$tags$retention_summary
```

These summaries help you reproduce and justify filtering decisions in manuscripts and reports.
