---
title: "proteoDA_v1.0_workflow"
author: "Byrum Lab"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{proteoDA v1.0 Workflow Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "2025-11-24"
---

```{r setupknit, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Installation

To install proteoDA v1.0 from GitHub (with dependencies and vignettes built):

```{r install, eval=FALSE}
devtools::install_github("ByrumLab/proteoDA",
                         dependencies = TRUE,
                         build_vignettes = TRUE,
                         force = TRUE)

```

## Load the package and view the available vignettes:

```{r load packages, echo=TRUE, eval=TRUE}
library(proteoDA)
# browseVignettes(package = "proteoDA")
```


# Input File Requirements and example

`proteoDA` ships with small DIA-style test datasets for quick evaluation. Three data matrix are required as input: 

1. protein intensity data - proteins are in rows and samples are in columns
2. sample metadata - links sample columns in the protein intensity to sample group information
3. protein annotation - includes `uniprot_id` required column, protein description, gene names, and any other protein information. 

*NOTE:* 
1. The protein intensity dataframe must include a column labeled `uniprot_id`. This becomes the rownames for the lists in R and must have unique values for each row. 
2. The sample metadata csv file should have a column labeled `sample` that matches the colnames of the samples in the protein intensity dataframe. 

```{r package_data}
input_data <- read.csv(system.file("extdata/DIA_data.csv.gz", package = "proteoDAstjude"))
sample_metadata <- read.csv(system.file("extdata/metafile.csv", package = "proteoDAstjude"))

head(input_data)
head(sample_metadata)

```

## Preparing the input files

Most of the protein data files generated from various database search engines will include the protein annotation along with the protein intensities for each sample in the same file. In order to create the `DAList` in `proteoDA`, we need to split the database search results into the protein annotation and intensity data. Only the columns needed for analysis should be included. 


```{r setup_anno_intensity}

annotation_data <- input_data[, 1:4]            # protein-level annotations
intensity_data  <- input_data[, 5:ncol(input_data)]  # quantitative sample columns

# Ensure metadata rownames match intensity column names
rownames(sample_metadata) <- sample_metadata$data_column_name

```

# Create a `DAList` Object

`DAList` is the central container class in proteoDA for holding intensity, metadata, and downstream analysis results.

```{r create_DAList}
raw <- DAList(
data      = intensity_data,
annotation= annotation_data,
metadata  = sample_metadata
)

```

# Preprocessing

## Filter Samples

The `filter_samples()` allows you to remove any outlier samples, pooled or QC samples not used in limma group comparisons. 

```{r filter_samples}
# samples with the group named "Pool" will be removed 
filtered_samples <- filter_samples(raw, group != "Pool")

```

## How proteoDA v1.0 deals with missing values 

Depending on your experimental design, you may wish to treat intensity values below the detection limit of your instrument as missing data (NA) or as a value of 0 intensity (protein is truly knocked out). Though this can be done manually before importing data into a DAList object, we also provide some simple utility functions to convert missing values to 0 and vice-verse. 

`missing_to_zero()` 
- converts missing values (NA by default, or other user-provided values) to 0 
- is mainly for downstream export/compatibility for tools that require a fully numeric matrix, not for model fitting, such as performing PCA analysis

`zero_to_missing()` 
- converts 0s to NA
- limma treats them as missing rather than true measurements.
- NA values are required for `filter_proteins_by_group()` and for the missing value heatmap

### How limma treats NA values
The way NA vs 0 is handled is exactly how limma does it where NA values in the expression matrix are treated as missing.

For each protein/feature, lmFit():
1. drops samples that are NA for that feature only, and
2. fits the linear model using only the remaining non-missing samples.

This means:
- Different proteins can be fit with different numbers of samples.
- If a whole group (e.g. all “cancer” samples) is NA for a protein, the corresponding design column / contrast becomes non-estimable and `limma` returns NA for logFC, t-statistic, p-value for that protein/contrast.
- `eBayes()` uses the available residual degrees of freedom per protein, so proteins with many missing values borrow information but still have appropriately smaller df.

### How limma treats zeros

- A value of 0 is just a number to limma.
- On the log2 scale, 0 means “intensity of 1”; on raw scale it means “exactly zero”.

- `lmFit()` and `eBayes()` assume these zeros are real observed values and

1. includes them in the mean and variance calculations,
2. allows them to contribute to fold-changes and t-statistics.

If zeros actually mean “not detected / missing” rather than a true measured abundance, treating them as real values can:

1. Pull group means strongly downward,
2. Inflate log2 fold-changes when one group has many zeros and the other doesn’t,
3. Artificially increase within-group variance for “mixed” features (some real intensities, some zeros),
4. Distort the moderated variance estimates in eBayes().

*NOTE:* NA = “this measurement is not available; ignore it for this protein’s model fit”. limma never tries to impute internally.

`proteoDA v1.0` does not include an imputation method; however, version 2.0 now includes imputation using the normal distribution as performed in the Perseus Software (Cox lab). 

*Recommendataion:* Convert explicit zeros to NA (recommended before filtering):

```{r zero_to_NA}
filtered_samples <- zero_to_missing(filtered_samples)
```

## Filter Proteins by Group

Filter features based on replicate presence (e.g., ≥4 values in ≥1 group):

This function is used to remove proteins from a DAList, filtering out proteins based on levels of missing values in the "data" data frame of the `DAList`. The `grouping_column` must be a column in the metadata of the `DAList`, which lists the group membership for each sample. The `min_reps` and `min_groups` arguments that determine the number of replicates/samples per group (`min_reps`) and number of groups (`min_groups`) in which a protein must have non-missing intensity values in order to be retained. This function assumes that all missing values are encoded as NA.

*Rule of Thumb* require at least 2/3 of replicates to have values in at least one sample group. 

```{r filter_proteins}
filtered_proteins <- filter_proteins_by_group(
filtered_samples,
min_reps        = 4,
min_groups      = 1,
grouping_column = "group"
)

```

# Normalization and QC

`write_norm_report()` saves a PDF report containing a variety of qualitative plots which give information about the performance of different normalization methods. The report is useful for choosing a normalization method to use for downstream analysis.

## Evaluating Normalization Methods with write_norm_report()

Proteomics intensity data almost always contain systematic, non-biological biases—arising from sample handling, instrument drift, batch effects, or other uncontrolled sources. As emphasized in the `proteiNorm` publication (Graw et al), selecting an appropriate normalization strategy is a critical preprocessing step as poor normalization can obscure true biological changes, inflate variance, or introduce artificial fold-changes.

To support this evaluation in `proteoDA`, the function `write_norm_report()` provides an automated, reproducible way to compare the performance of eight widely used normalization methods. Internally, this wrapper runs the same QC, normalization, and diagnostic steps implemented in the `proteiNorm` framework, but produces a static PDF or HTML report suitable for documentation and publication.

### What normalization methods are evaluated?

`write_norm_report()` applies the following methods to the protein abundance matrix

1. Log2 transformation (log2)
2. Median normalization (median)
3. Mean normalization (mean)
4. Variance Stabilizing Normalization (vsn)
5. Quantile normalization (quantile)
6. Cyclic loess normalization (cycloess)
7. Robust linear regression (rlr)
8. Global intensity scaling (gi)

These methods represent a combination of microarray-derived normalization strategies and proteomics-specific regressions/scaling approaches, many of which consistently rank among the top performers in systematic evaluations of label-free proteomics data.

### What diagnostics does the report generate?

Following the `proteiNorm` workflow, the `proteoDA` report evaluates each method with a consistent QC panel:

1. *Total intensity boxplots*: Checks whether global sample intensities become comparable after normalization.
2. *PCA plots*: Visualizes grouping structure, batch effects, and separation of biological conditions.
3. *Missing value heatmap*: Displays patterns of missingness (MAR vs MNAR) as clustered heatmap.
4. *Intragroup variation metrics*

  For each normalization method, the report quantifies:
- PCV (pooled coefficient of variation)
- PEV (pooled estimate of variance)
- PMAD (pooled median absolute deviation)

These metrics are key evaluation criteria in `proteiNorm` because lower intragroup variation indicates a more stable, better-performing normalization method.

5. *Correlation metrics*: 

- Boxplots of within-group correlations
- Full Pearson correlation heatmaps 

Higher intragroup correlation suggests that biological replicates align after normalization.

6. *Log2 ratio distributions*: Distribution of all pairwise log2 fold-changes between sample groups, expected to center around zero unless strong directed regulation is known, such as a 2-fold spike in study.

### How to interpret the results

The goal is to choose the normalization method that:
- Minimizes within-group variation (PCV/PEV/PMAD)
- Maximizes correlation among replicates
- Produces centered log2-ratio distributions
- Yields PCA plots with expected grouping and minimal batch effects
- Shows stable, comparable total intensities

As demonstrated in the `proteiNorm` paper’s case studies (mouse, breast cancer, spike-in datasets) , different experiments may favor different methods—VSN and cyclic loess often rank highly, but no single method is optimal for all designs. write_norm_report() makes these comparisons reproducible for any proteoDA workflow.

*Reference:* Graw, S., Tang, J., Zafar, M. K., Byrd, A. K., Bolden, C., Peterson, E. C., & Byrum, S. D. (2020). proteiNorm – A User-Friendly Tool for Normalization and Analysis of TMT and Label-Free Protein Quantification. ACS Omega, 5(40), 25625-25633. https://doi.org/10.1021/acsomega.0c02564


```{r normalization_QC}
write_norm_report(
filtered_proteins,
grouping_column = "group",
output_dir = "01_QC_report_v1.0",
filename = "normalization.pdf",
overwrite = TRUE
)

# Available methods: "log2", "median", "mean", "vsn", "quantile", "cycloess", "rlr", "gi"

normalized <- normalize_data(filtered_proteins, norm_method = "cycloess")

```

### Create a QC report including PCA and clustering:

```{r qc_report}
write_qc_report(
  normalized,
  color_column = "group",
  output_dir = "01_QC_report_v1.0",
  filename = "QC_report.pdf",
  overwrite = TRUE,
  top_proteins = 500,
  standardize  = TRUE,
  pca_axes     = c(1, 2),
  dist_metric  = "euclidean",
  clust_method = "complete"
)

```

## 6. Differential Analysis (No-Intercept Model)


## 6.1 Add limma design and contrasts. 

For standard group comparisons, use a no-intercept model: `~ 0 + group`

Contrasts can be added using either the `contrasts_vector` or the `contrasts_file` option. 
When using the file option, save the contrasts in column A as a .csv file. 

[Provide example table here]

```{r add_design_contrasts}
no_intercept <- add_design(normalized, design_formula = ~0 + group)

# Define contrasts of interest
no_intercept <- add_contrasts(
no_intercept,
contrasts_vector = "cancer_vs_normal = cancer - normal"
)

no_intercept$design$design_matrix
```

## 6.2 Fit the linear model

When you call fit_limma_model(no_intercept) in proteoDA, under the hood it’s doing:
fit <- limma::lmFit(exprs, design)
fit <- limma::eBayes(fit)



```{r model_fit}
fit <- fit_limma_model(no_intercept)
```

## 6.3 Extract the differential analysis results: 

The `pval_thresh` and `lfc_thresh` parameters will define how significant features are defined and highlighted in plots and tables. 

```{r extract_DA_results}
results <- extract_DA_results(
fit,
pval_thresh      = 0.05,
lfc_thresh       = 1,
adj_method       = "BH",
extract_intercept = FALSE
)

```

# 7. Reporting Results

## Write statistical tables (CSV, Excel, and summary):

```{r write_limma_tables}
# in version 2.0, DA_table_cols must be defined in order to write the contrasts sheets. 
DA_table_cols <- c("uniprot_id","Accession.Number","Protein.Description")

write_limma_tables(
results,
output_dir        = "02_DA_results_v1.0",
overwrite         = TRUE,
add_filter        = TRUE,
contrasts_subdir  = NULL,
combined_file_csv = NULL,
spreadsheet_xlsx  = NULL
)
```

## Generate interactive and static volcano plots for each contrast:

```{r write_limma_plots}

write_limma_plots(
results,
grouping_column = "group",
table_columns   = c("uniprot_id", "protein_info"),
title_column    = "uniprot_id",
output_dir      = "02_DA_results_v1.0",
tmp_subdir      = "tmp",
height          = 1000,
width           = 1000,
overwrite       = TRUE
)

```

# 8. Summary

## This workflow demonstrates the standard proteoDA v1.0 pipeline:

1. Create a `DAList` from raw data and metadata
2. Filter samples and proteins
3. Normalize and QC
4. Define the limma model with `add_design()` and `add_contrasts()`
5. Fit the model with `fit_limma_model()`
6. Extract results with `extract_DA_results()`
7. Generate reports with `write_limma_tables()` and `write_limma_plots()`

This basic structure extends easily to factorial, paired, or blocked designs using the same core functions (add_design, add_contrasts, and fit_limma_model).

# 9. Advanced: 2×2 Factorial Design (Intercept Model with Interaction)

This example demonstrates a 2×2 factorial design with interaction using the **core v1.0 API** only.  
We assume your metadata has two factors:

- `cell` with levels `WT`, `HET` (genotype)  
- `treat` with levels `Sham`, `Gy` (treatment)

> **Note:** `proteoDA` simplifies design-matrix column names (e.g., `Intercept`, `HET`, `Gy`, `HET.treatGy`) so contrasts and plot labels are cleaner.

## 9.1 set sample group reference levels 

```{r intercept_ref_levels, echo=TRUE, eval=FALSE}
## ---- nine_setup_factors, echo=TRUE ------------------------------------------
# Ensure the factors exist in the metadata and set reference levels
stopifnot(all(c("cell", "treat") %in% names(sample_metadata)))

normalized$metadata$cell  <- factor(normalized$metadata$cell,  levels = c("WT","HET"))
normalized$metadata$treat <- factor(normalized$metadata$treat, levels = c("Sham","Gy"))

```

## 9.2 setup contrasts example

```{r Int_contrasts_example}

## ---- nine_show_contrasts_table, echo=FALSE, message=FALSE -------------------
# Render a small table of the contrasts we will use (for the vignette)
contrasts_df <- data.frame(
  Label = c(
    "WT_Gy_vs_HET_Gy",
    "WT_Sham_vs_HET_Sham",
    "TreatEffect_WT",
    "TreatEffect_HET",
    "Diff_in_effect_WT_HET",
    "Avg_Treat_Gy_vs_Sham_unscaled"
  ),
  Definition = c(
    "(Intercept + Gy) - (Intercept + HET + Gy + HET.treatGy)",
    "Intercept - (Intercept + HET)",
    "Gy",
    "Gy + HET.treatGy",
    "Gy - (Gy + HET.treatGy)",
    "(Intercept + Gy) + (Intercept + HET + Gy + HET.treatGy) - (Intercept) - (Intercept + HET)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(contrasts_df, caption = "Example contrasts for 2×2 factorial (intercept model)")

```

### Save the exact same contrasts to a CSV that add_contrasts() can read
### (a simple one-contrast-per-line plain text file also works; here we use writeLines for clarity)

```{r int_write_contrasts}
## ---- nine_write_contrasts_csv, echo=TRUE ------------------------------------
# Save the exact same contrasts to a CSV that add_contrasts() can read
# (a simple one-contrast-per-line plain text file also works; here we use writeLines for clarity)
fac_con_lines <- c(
  "WT_Gy_vs_HET_Gy = (Intercept + Gy) - (Intercept + HET + Gy + HET.treatGy)",
  "WT_Sham_vs_HET_Sham = Intercept - (Intercept + HET)",
  "TreatEffect_WT = Gy",
  "TreatEffect_HET = Gy + HET.treatGy",
  "Diff_in_effect_WT_HET = Gy - (Gy + HET.treatGy)",
  "Avg_Treat_Gy_vs_Sham_unscaled = (Intercept + Gy) + (Intercept + HET + Gy + HET.treatGy) - (Intercept) - (Intercept + HET)"
)
writeLines(fac_con_lines, "contrasts_factorial_intercept.csv")

```

## 9.3 Add design and contrasts for an interaction model (two cell types and two treatments). 

### The factorial approach formally tests whether treatment effects differ between cell lines.
###	Two-factor encoding: ~ treat1 * treat2 includes a main effect and an interaction term; the : column quantifies interaction (synergy or repression).

```{r int_add_design_and_contrasts, echo=TRUE, eval=FALSE}
# Add factorial design with interaction (intercept model)
fac <- add_design(
  normalized,
  design_formula = ~ cell * treat
)

# Add contrasts from the CSV we just wrote
fac <- add_contrasts(
  fac,
  contrasts_file = "contrasts_factorial_intercept.csv"
)

# (Optional sanity check)
colnames(fac$design$design_matrix)
# Expected: "Intercept", "HET", "Gy", "HET.treatGy"
```

## 9.4 Fit and extract results

```{r int_fit_results, eval=FALSE}
## ---- nine_fit_and_extract, echo=TRUE ----------------------------------------
# Fit and extract (v1.0 core functions)
fit_fac <- fit_limma_model(fac)

results_fac <- extract_DA_results(
  fit_fac,
  pval_thresh       = 0.05,
  lfc_thresh        = 1,
  adj_method        = "BH",
  extract_intercept = FALSE   # keep FALSE: we are interested in contrasts, not the Intercept coefficient
)

# Write tables and plots for the factorial contrasts
write_limma_tables(
  results_fac,
  output_dir = "03_DA_factorial",
  overwrite  = TRUE,
  add_filter = TRUE
)

write_limma_plots(
  results_fac,
  grouping_column = "group",                 # or another metadata column for color
  table_columns   = c("uniprot_id","protein_info"),
  title_column    = "uniprot_id",
  output_dir      = "03_DA_factorial",
  tmp_subdir      = "tmp",
  height          = 1000,
  width           = 1000
)


```

## 9.5 Interpretation of the Interaction model `~ cell * treat`

**Interpretation tips**

- `TreatEffect_WT` is the Gy vs Sham effect **within WT** (reference genotype).  
- `TreatEffect_HET` is the Gy vs Sham effect **within HET** (WT treatment effect + interaction).  
- `Diff_in_effect_WT_HET` compares treatment effects between genotypes (the interaction contrast).  
- `Avg_Treat_Gy_vs_Sham_unscaled` is the **sum** of the two genotype-specific treatment effects.  
  If you need the literal average, divide the logFC by 2 after the fact in a downstream step (see proteoDA version 2.0 tutorial)


# 10. Paired / Blocking Designs (v1.0 Functions)

This section illustrates two approaches to handle non-independent samples in proteoDA v1.0:

- **Paired fixed-effect model**: explicitly include a `pair_id` term in the design (~ 0 + group + pair_id).  
- **Blocking with `duplicateCorrelation`**: set `DAList$design$random_factor` to the blocking column name before calling `fit_limma_model()`.  
  The function then estimates the intra-block correlation internally and applies it to the limma fit.

> ✅ Use the fixed-effect approach when the number of pairs is modest and you want strict within-pair comparisons.  
> ✅ Use `duplicateCorrelation` when you have many features and repeated measures (efficient and robust for omics).

## 10.1 Paired (fixed-effect) model

Design: `~ 0 + group + pair_id`
Each pair gets its own coefficient.
contrasts compare groups within pairs.

```{r 10_paired_effect_setup, echo=TRUE, eval=FALSE}
## ---- ten_paired_fixed_effect_setup, echo=TRUE --------------------------------
# Requirements in metadata:
#  - a grouping variable used in your comparison, e.g. "group" with levels {Pre, Post}
#  - a pairing variable, e.g. "pair_id" (or "subject"), repeated for matched samples

# Metadata must contain both "group" and "pair_id"
stopifnot(all(c("group","pair_id") %in% names(normalized$metadata)))

# Ensure factors (and set reference levels if needed)
normalized$metadata$group   <- factor(normalized$metadata$group)         # e.g., c("Pre","Post")
normalized$metadata$pair_id <- factor(normalized$metadata$pair_id)

# Add paired fixed-effect design
paired_fe <- add_design(
  normalized,
  design_formula = ~ 0 + group + pair_id
)

# Contrasts: Post vs Pre (labels show up in plots)
paired_fe <- add_contrasts(
  paired_fe,
  contrasts_vector = "Post_vs_Pre = Post - Pre"
)

# Fit and extract
fit_paired_fe <- fit_limma_model(paired_fe)

res_paired_fe <- extract_DA_results(
  fit_paired_fe,
  pval_thresh       = 0.05,
  lfc_thresh        = 0.5,
  adj_method        = "BH",
  extract_intercept = FALSE
)

# Write outputs
write_limma_tables(
  res_paired_fe,
  output_dir = "10A_Paired_FixedEffect",
  overwrite  = TRUE,
  add_filter = TRUE
)

write_limma_plots(
  res_paired_fe,
  grouping_column = "group",
  table_columns   = c("uniprot_id","protein_info"),
  title_column    = "uniprot_id",
  output_dir      = "10A_Paired_FixedEffect",
  tmp_subdir      = "tmp",
  height          = 1000,
  width           = 1000
)


```

### Notes

1. This absorbs baseline differences per pair via a fixed effect (`pair_id` defined in sample_metadata.csv).
2. Use when every pair has both conditions (e.g., Pre/Post) and the number of pairs is not huge.

## 10.2 Blocking with internal duplicateCorrelation

Design: `~ 0 + group`
Random factor: assign the blocking column name to `DAList$design$random_factor`.

`fit_limma_model()` will:

1. Estimate the consensus correlation via `duplicateCorrelation()`
2. Inform the user of the estimated value
3. Apply it automatically during the model fit. 

limma estimates an intra-block correlation and uses it in the model fit—computationally efficient and close in spirit to a random intercept.


```{r 10_blocking-dupCor_setup, echo=TRUE, eval=FALSE}
## ---- ten_blocking_dupCor_setup, echo=TRUE -----------------------------------
# Requirements in metadata:
#  - a grouping variable (e.g. "group" with levels {Pre, Post})
#  - a blocking variable (e.g. "pair_id" or "subject"), repeated for correlated samples

stopifnot(all(c("group","pair_id") %in% names(normalized$metadata)))

# Factor coding (reference choice doesn't matter for no-intercept designs)
normalized$metadata$group   <- factor(normalized$metadata$group)
normalized$metadata$pair_id <- factor(normalized$metadata$pair_id)

# Add simple group-means design (no intercept)
blocked <- add_design(
  normalized,
  design_formula = ~ 0 + group
)

# Contrast: Post vs Pre
blocked <- add_contrasts(
  blocked,
  contrasts_vector = "Post_vs_Pre = Post - Pre"
)

# Specify the blocking variable for duplicateCorrelation
blocked$design$random_factor <- "pair_id"

# Fit the model; duplicateCorrelation will run automatically
fit_blocked <- fit_limma_model(blocked)

res_blocked <- extract_DA_results(
  fit_blocked,
  pval_thresh       = 0.05,
  lfc_thresh        = 0.5,
  adj_method        = "BH",
  extract_intercept = FALSE
)

# Write outputs
write_limma_tables(
  res_blocked,
  output_dir = "10B_Blocked_dupCor",
  overwrite  = TRUE,
  add_filter = TRUE
)

write_limma_plots(
  res_blocked,
  grouping_column = "group",
  table_columns   = c("uniprot_id","protein_info"),
  title_column    = "uniprot_id",
  output_dir      = "10B_Blocked_dupCor",
  tmp_subdir      = "tmp",
  height          = 1000,
  width           = 1000
)

```

### Interpretation

1. `duplicateCorrelation()` estimates an average within-pair/within-subject correlation and passes it to lmFit() so the residual variance accounts for dependence.

2. Great for repeated measures and many features (omics), where a full mixed model would be heavy.

3. `fit_limma_model()` prints an estimated intra-block correlation (e.g., 0.12).
If it’s low (< 0.05), the function issues a yellow message suggesting that a simple model may suffice.

4. This approach is efficient for large omics datasets and conceptually similar to a random-intercept mixed model.

5. It provides most of the benefit of modeling correlation within subjects without the heavy computation of lme4 or dream.

### Choosing between the two

1. Use **paired fixed-effect** when each pair is complete and you want strict within-pair comparisons (each pair gets a coefficient).

2. Use **blocking** (duplicateCorrelation) for scalability and when you expect a common within-block correlation across features; conceptually similar to a **random intercept**.

# 11. Diagnostics for models 

## Checks full rank, condition number, non-estimable contrasts, and (if using blocking) prints the duplicateCorrelation estimate. It works for either path:

1. paired_fe (paired fixed-effect model)
2. blocked (blocking with random_factor)

```{r 11_diagnostics_models, echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}

## ---- diagnostics_model_checks, echo=TRUE, message=FALSE, warning=FALSE -------
# Helper to print a clean header
.hdr <- function(x) cat("\n", paste0("## ", x), "\n", sep="")

# Choose which object to diagnose (set one that exists)
obj <- if (exists("blocked")) blocked else if (exists("paired_fe")) paired_fe else NULL
stopifnot(!is.null(obj))

X <- obj$design$design_matrix
C <- obj$design$contrast_matrix

.hdr("Design matrix summary")
cat("Dimensions (n x p):", paste(dim(X), collapse=" x "), "\n")
cat("Column names:\n"); print(colnames(X))

# Full-rank / rank deficiency check
qrX   <- qr(X)
rankX <- qrX$rank
p     <- ncol(X)
cat("QR rank:", rankX, "  |  Columns:", p, "\n")
if (rankX < p) {
  cat("WARNING: design is NOT full rank. Check factor coding / redundant columns.\n")
} else {
  cat("OK: design is full rank.\n")
}

# Condition number as a rough collinearity diagnostic
k <- tryCatch(kappa(X), error = function(e) NA_real_)
cat("Condition number (kappa):", if (is.na(k)) "NA" else sprintf("%.2f", k), "\n")
if (!is.na(k) && k > 1e6) cat("NOTE: Very high kappa suggests near-collinearity.\n")

# Contrast estimability
.hdr("Contrast estimability (limma::nonEstimable)")
if (!is.null(C)) {
  ne <- limma::nonEstimable(C, X)
  if (is.null(ne)) {
    cat("All contrasts appear estimable given the design.\n")
  } else {
    cat("Non-estimable (or suspect) contrasts detected:\n")
    print(ne)
  }
} else {
  cat("No contrast matrix present.\n")
}

# If using blocking via random_factor, show duplicateCorrelation estimate
.hdr("Blocking / duplicateCorrelation diagnostic")
if (!is.null(obj$design$random_factor)) {
  blk_col <- obj$design$random_factor
  stopifnot(blk_col %in% colnames(obj$metadata))
  block_vec <- obj$metadata[, blk_col, drop = TRUE]
  dc <- limma::duplicateCorrelation(object = obj$data, design = X, block = block_vec)
  cat("Estimated intra-block correlation (duplicateCorrelation):",
      sprintf("%.3f", dc$consensus.correlation), "\n")
} else {
  cat("No random_factor set; duplicateCorrelation not applied.\n")
}

# Quick peek at coefficients and residual df if already fitted
.hdr("Fit overview (if available)")
if (!is.null(obj$eBayes_fit)) {
  efit <- obj$eBayes_fit
  cat("Coefficients matrix dims:", paste(dim(efit$coefficients), collapse=" x "), "\n")
  if (!is.null(efit$df.residual)) cat("Residual df (median across features):",
                                      stats::median(efit$df.residual, na.rm = TRUE), "\n")
  # Show first few coefficient names
  cat("Coefficient names:\n"); print(colnames(efit$coefficients))
} else {
  cat("Model not fitted yet (no eBayes_fit found in object).\n")
}


```

## What it does:

1. Full rank: uses `qr()` to compare rank vs. number of columns; warns if deficient.
2. Collinearity: reports `kappa(X)` (large values ~ near-collinearity).
3. Estimability: uses `limma::nonEstimable()` to flag contrasts that aren’t estimable under the design.
4. Blocking: if `obj$design$random_factor` is set, re-computes and prints the duplicateCorrelation consensus correlation (purely diagnostic).
5. Fit summary: if you’ve already run `fit_limma_model()`, prints coefficient dimensions, residual df, and coefficient names.